Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2020-12-31 SIR_manuscript.tex
working on intro

The observed dynamics of the COVID-19 epidemic is dirven by the integrated processes including both epidemiological processes (infection, recovery, etc.) and testing processes. In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the control of the epidemic.

Spereating the epidemiological processes from testing processes through a mathematical modeling excercise provides a tool to investigate the effect of the testing on the epidemic dynamics. Although, this effect is intuitively through isolation and quarantine practices, the sensitivity of the epidemic dynamics to these practices needs to be studied further. During COVID-19 epidemic, the isolation has applied to people who have been tested. Thus, the epidemic dynamics will depend on testing strategies which determines how tests are distributed across epidemiological compartments, or for a given testing intensity how many people in each compartment get tested. Specifically, in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be distributed equally across all compartments.

that the epi effects will depend on \emph{who gets tested}, i.e. how tests are distributed across compartments, or for a given testing intensity how many people in each epidemiological compartment get tested. Then you can go through some examples, i.e. start by saying that in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be weighted equally across all compartments \ldots}

% One of the main challenges in understanding COVID-19 dynamics is to unravel information available from data. The daily case count is a highly accessible form of data where observations are driven by two clusters of processes: (i) epidemic processes and (ii) testing processes. The challenge is then to separate these processes and understand how they interact. Simple mathematical models have a long history in providing useful insights facing with real epidemiological challenges \citep{ross1911prevention} [maybe add influenza work, ideas?]. We developed a model that incorporates mechanistic processes of epidemic processes and testing. We used the model to study the potential effect of testing strategies on disease dynamics.

% \jd{There is a gap here between unravelling data and what we actually do.}
% \bmb{As JD suggests, the focus of the paper is not on unraveling data (although that is something we \emph{could} do, and are trying to do, with this modeling framework as well). I would say that the ``bullet points'' to start the paper are (1) the processes driving the observed dynamics of the COVID-19 epidemic include both epidemiological processes (infection, recovery, etc.) and testing processes (``processes'' is repeated here, could rephrase); (2) in addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine; (3) your last sentence in the \P\ above.}

\jd{This \P\ is confusing. The topic sentence seems to be about kinds of tests, and then there is a jump to sampling strategies.}

\bmb{It's fine for this \P\ to be about testing, but the topic sentence still doesn't say what the paragraph is trying to do. You should probably give the general idea of how isolation and quarantine affect epi dynamics (i.e. individuals who in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate), and say that your goal is to figure out the effects of isolation and testing: when can isolation and testing control an epidemic? You might even want to do this in the \emph{previous} paragraph. Then you can say that since isolation applies to people who have been tested, that the epi effects will depend on \emph{who gets tested}, i.e. how tests are distributed across compartments, or for a given testing intensity how many people in each epidemiological compartment get tested. Then you can go through some examples, i.e. start by saying that in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be weighted equally across all compartments \ldots}

@@

Further, the testing strategies can be grouped into two categories including random testing across the whole population and Non-random/biased testing. Random testing is the simplest form of testing strategy and is typically carried out for the purpose of surveillance. It is notable that, by random testing we mainly refer to PCR test given that there is almost negligible virological testing for surveillance purposes. Furthermore, all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are non-random testing strategies with a different bias. Specifically, when the purpose is "(i) diagnosis and treatment" or "(iv) mitigation and control of disease spread", the bias is towards infectious individuals more strongly for (i) than (iv). When people are tested for "(ii) screening", the bias is towards the group-specific characteristics [I am making this up for now!]. For example, people who are getting on flights may/will be more mobile than people awaiting for a surgery and are probably going to stay in long-term care homes and consequently more isolated to begin with.

-Examples of random versus biased testing strategies;
Each of these strategies has been used as a tool to study COVID-19; see (refs) for group (i) and (refs) for group (ii). The advantage of non-random testing comparing to random testing has been investigated specifically during the COVID-19 pandemic (refs).

>>>>

The observed dynamics of the COVID-19 epidemic is dirven by the underlying processes including both epidemiological processes (infection, recovery, etc.) and testing processes. In addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine. Specifically, individuals who are in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate. We developed a model that incorporates mechanistic epidemic processes and testing to explore the effects of testing and isolation on the epidemic dynamics, and providing insights for the epidemic management/control.

Spereating the epidemiological processes from testing processes in a mechanistic modeling framework provides a tool to investigate the effect of the testing on the epidemic dynamics. Intutively, this effect is through isolation and quarantine practices. The informative/useful part is the sensitivity of the epidemic dynamics to testing, isolation and quarantine. Specifically, during COVID-19 epidemic, the isolation has applied to people who have been tested. Thus, the epidemic dynamics will depend on testing processes. The latter determines how tests are distributed across epidemiological compartments, or for a given testing intensity how many people in each compartment get tested. Specifically, in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be weighted equally across all compartments. However, in the case where tests are targeted certain clusters of the population (as might happen, for example, for diagnosis and treatment, screening people for access to things like flights, surgery, long-term-care facilities, etc., and for mitigation and control of the disease spread) the tests will be weighted differently, where the greater the testing weight for a compartment, the more likely people in that compartment get tested. \todo{edit the last sentence, it is very long!}

% \bmb{As JD suggests, the focus of the paper is not on unraveling data (although that is something we \emph{could} do, and are trying to do, with this modeling framework as well). I would say that the ``bullet points'' to start the paper are (1) the processes driving the observed dynamics of the COVID-19 epidemic include both epidemiological processes (infection, recovery, etc.) and testing processes (``processes'' is repeated here, could rephrase); (2) in addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine; (3) your last sentence in the \P\ above.}

@@

It is notable that, by random testing we mainly refer to the PCR test given that there is almost negligible virological testing for surveillance purposes. Furthermore, all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are targeted testing strategies with a different bias. Specifically, when the purpose is "(i) diagnosis and treatment" or "(iv) mitigation and control of disease spread", the bias is towards infectious individuals more strongly for (i) than (iv). When people are tested for "(ii) screening", the bias is towards the group-specific characteristics [I am making this up for now!]. For example, people who are getting on flights may/will be more mobile than people awaiting for a surgery and are probably going to stay in long-term care homes and consequently more isolated to begin with.
